Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2024-10-24 index.qmd
section:issues; process:write

<!-- 2. But what are the problems of DCJ as a method? -->
Nevertheless, despite the growing number of studies on CJ, unsystematic and fragmented research approaches in the literature have left several critical issues unaddressed. This research primarily focuses on three: the apparent disconnect between CJ's structural and measurement model, the over-reliance on the assumptions of Thurstone's Case 5 [-@Thurstone_1927] in CJ's measurement model, and the unclear role and impact of comparison algorithms and the number of comparisons on the method's reliability and validity. The following sections will discuss each of these issues in detail, followed by the introduction of a theoretical model and its statistical translation, which aim to address all three concerns simultaneously.

# The three critical issues in CJ literature {#sec-theory-issues}

@@

<!-- 1. The BTL model is CJ's measurement model -->
In a typical CJ study, the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959] serves as the measurement model. This model specifies how latent variables are estimated from observed variables [@Everitt_et_al_2010]. In CJ, multiple judges engage in several rounds of pairwise comparisons to assess the relative manifestation of a trait between two stimuli. Each comparison generates a dichotomous outcome, indicating which stimulus is perceived to exhibit a higher degree of the trait. The BTL model then uses these observed outcomes to estimate scores that represent the latent trait of interest [@Pollitt_2012a; @Pollitt_2012b; @Whitehouse_2012; @Jones_2015; @vanDaal_et_al_2016; @Lesterhuis_2018; @Boonen_et_al_2020; @Bouwer_et_al_2023].

<!-- 2. But there is no structural model, only post estimation analyses -->
Moreover, researchers often use these BTL-generated scores or their transformations separately, conducting further analyses or hypothesis testing. The CJ literature highlights how these scores have been employed to identify 'misfit' judges and stimuli [@Pollitt_2012b; @vanDaal_et_al_2017; @Goossens_et_al_2018], detect biases in judges' ratings [@Pollitt_et_al_2003; @Pollitt_2012b], calculate correlations with other scoring methods [@Goossens_et_al_2018; @Bouwer_et_al_2023], or test hypotheses related to the trait of interest [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].

<!-- 3. But what is the problem with this? and what is the solution? -->
However, the statistical literature cautions against using estimated scores for separate analyses and hypothesis testing. A key consideration is that BTL-generated scores are parameter estimates that inherently carry uncertainty. Ignoring this uncertainty when conducting separate analyses and test can inflate their precision and statistical power. This, in turn, increases the risk of committing a type I error, which is when a null hypothesis is incorrectly rejected [@McElreath_2020]. To mitigate this risk, principles from Structural Equation Modeling (SEM) [@Hoyle_et_al_2023; @Kline_et_al_2023] recommend conducting these analyses within a structural model that accounts for both the scores and their uncertainties, rather than treating them separately. Thus, an integrated approach combining CJ's structural and measurement models can offer significant advantages.

@@

<!-- (a) related to Case 3, it assumes the correlation between stimuli is zero, this translates into the cancellation of judges effects by mean of opposing and equally weigthed 'mood' and 'simultaneous contrasts' effects. This is demonstrated using the additivive nature of the logit scale, which helps to cancel 'bias' judges effects. -->
<!-- This is the case of confounding, and @Andrich_1978; @Wainer_et_al_1978 already showed that if judges effects are parametrized, it gets eliminated experimentally in the paired comparison method.  -->

>>>>

<!-- 2. But what are the problems of DCJ as a method? -->
Nevertheless, despite the growing number of studies on CJ, unsystematic and fragmented research approaches in the literature have left several critical issues unaddressed. This research primarily focuses on three: the apparent disconnect between CJ's structural and measurement model, the over-reliance on the assumptions of Thurstone's Case 5 [-@Thurstone_1927] in CJ's measurement model, and the unclear role and impact of comparison algorithms on the method's reliability and validity. The following sections will discuss each of these issues in detail, followed by the introduction of a theoretical model and its statistical translation, which aim to address all three concerns simultaneously.

# Three critical issues in CJ literature {#sec-theory-issues}

@@

<!-- 1. The BTL model is CJ's measurement model -->
In a typical CJ study, the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959] serves as the measurement model [@Andrich_1978; @Bramley_2008]. This model specifies how latent variables are estimated from manifest variables [@Everitt_et_al_2010]. In CJ, multiple judges engage in several rounds of pairwise comparisons to assess the relative manifestation of a trait between two stimuli. Each comparison generates a dichotomous outcome, indicating which stimulus is perceived to exhibit a higher degree of the trait. The BTL model then uses these observed outcomes to estimate scores that represent the latent trait of interest [@Pollitt_2012a; @Pollitt_2012b; @Whitehouse_2012; @Jones_2015; @vanDaal_et_al_2016; @Lesterhuis_2018; @Boonen_et_al_2020; @Bouwer_et_al_2023].

<!-- 2. But there is no structural model, only post estimation analyses -->
Moreover, researchers often use these BTL-generated scores or their transformations separately, conducting further analyses or hypothesis testing. The CJ literature shows how these scores have been employed to identify 'misfit' judges and stimuli [@Pollitt_2012b; @vanDaal_et_al_2017; @Goossens_et_al_2018], detect biases in judges' ratings [@Pollitt_et_al_2003; @Pollitt_2012b], calculate correlations with other scoring methods [@Goossens_et_al_2018; @Bouwer_et_al_2023], or test hypotheses related to the trait of interest [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].

<!-- 3. But what is the problem with this? and what is the solution? -->
However, the statistical literature cautions against using estimated scores for separate analyses and tests. A key consideration is that BTL-generated scores are parameter estimates that inherently carry uncertainty. Ignoring this uncertainty when conducting separate analyses and tests can inflate their precision and statistical power. This, in turn, increases the risk of committing a type I error [@McElreath_2020], which is when a null hypothesis is incorrectly rejected [@Everitt_et_al_2010]. To mitigate this risk, principles from Structural Equation Modeling (SEM) [@Hoyle_et_al_2023; @Kline_et_al_2023] and Item Response Theory (IRT) [@deAyala_2009; @Fox_2010; @vanderLinden_et_al_2017] recommend conducting these analyses and tests within a structural model that accounts for both the scores and their uncertainties, rather than treating them separately. Thus, an integrated approach combining CJ's structural and measurement models can offer significant advantages.

@@

<!-- (a) related to Case 3, it assumes the correlation between stimuli is zero, this translates into the cancellation of judges effects by mean of opposing and equally weigthed 'mood' and 'simultaneous contrasts' effects. This is demonstrated using the additivive nature of the logit scale, which helps to cancel 'bias' judges effects. -->

<!-- This is the case of confounding, and @Andrich_1978; @Wainer_et_al_1978 already showed that if judges effects are parametrized, it gets eliminated experimentally in the paired comparison method.  -->
